---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 3 - Due date 02/01/24"
author: "Cara Kuuskvere"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r Load Packages}
#Load/install required package here
library(forecast)
library(tseries)
library(Kendall)
library(readr) 
library(ggplot2)
library(cowplot)
```

```{r Import Data}
#Importing data set
raw_energy_data <- read.csv(
  "Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.csv",
  header=FALSE,skip=12)
```

```{r Clean Data}
#trim the table
energy_data <- raw_energy_data[,4:6] #want all rows
                                 #all columns from 4 to 6
n_energy_sources <- ncol(energy_data) #number of variables
n_obs <- nrow(energy_data) #number observations

#Adding column names
colnames(energy_data)=c("Total Biomass Energy Production (Trillion Btu)",
                            "Total Renewable Energy Production (Trillion Btu)",
                            "Hydroelectric Power Consumption (Trillion Btu)")

head(energy_data)
```

```{r Transforming the data into time series object}
ts_energy_data <- ts(energy_data,start=c(1973,1),frequency = 12)
# frequency is 12 because it's monthly data, and months repeat every 12 entries!
# starts in year 1973
head(ts_energy_data)
```

##Trend Component

### Q1
For each time series, i.e., Renewable Energy Production and Hydroelectric Consumption create three plots: one with time series, one with the ACF and with the PACF. You may use the some code form A2, but I want all the three plots side by side as in a grid.
(Hint: use function `plot_grid()` from the `cowplot` package)

```{r Time Series Plots}
biomass_ts_plt <- autoplot(ts_energy_data[,1]) +
  labs(x="Year",y= "TBtu",
    title = "Biomass TS") 
#biomass_ts_plt

RE_ts_plt <- autoplot(ts_energy_data[,2]) +
  labs(x="Year",y= "TBtu",
  title = "Renewable Energy TS")
#RE_ts_plt

hydro_ts_plt <- autoplot(ts_energy_data[,3]) +
  labs(x="Year",y= "TBtu",
    title = "Hydroelectric TS")
#hydro_ts_plt

```

```{r Acf Plots}
biomass_acf <- ggAcf(ts_energy_data[,1],lag.max=40)+
  labs(title="Biomass ACF")
#biomass_acf
hydro_acf <- ggAcf(ts_energy_data[,3],lag.max=40)+
  labs(title="Hydro ACF")
#hydro_acf
RE_acf <- ggAcf(ts_energy_data[,2],lag.max=40)+
  labs(title="RE ACF")
#RE_acf
```

```{r Pacf Plots}
biomass_pacf <- ggPacf(ts_energy_data[,1],lag.max=40)+
  labs(title="Biomass PACF")
#biomass_pacf
hydro_pacf <- ggPacf(ts_energy_data[,3],lag.max=40)+
  labs(title="Hydro PACF")
#hydro_pacf
RE_pacf <- ggPacf(ts_energy_data[,2],lag.max=40)+
  labs(title="RE PACF")
#RE_pacf
```

```{r Putting them all together }
#plot_grid(biomass_ts_plt,biomass_acf,biomass_pacf, align="h",label_size=8)
#plot_grid(hydro_ts_plt,hydro_acf,hydro_pacf, align = "h", label_size=8)
#plot_grid(RE_ts_plt,RE_acf,RE_pacf,align="h",label_size=8)
plot_grid(biomass_ts_plt,biomass_acf,biomass_pacf,
          hydro_ts_plt,hydro_acf,hydro_pacf,
          RE_ts_plt,RE_acf,RE_pacf,
          ncol=3,align="h")
```


### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

Biomass and renewable energy production appear to have an increasing somewhat linear trend over time. The hydroelectric power consumption appears to have a strong seasonal component, which makes it hard to see if there is a small trend component. 

### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.
 
```{r}
t <- 1:n_obs
biomass_linear_trend <- lm(energy_data[,1]~t)
RE_linear_trend <- lm(energy_data[,2]~t)
hydro_linear_trend <- lm(energy_data[,3]~t)

summary(biomass_linear_trend)
summary(RE_linear_trend)
summary(hydro_linear_trend)
```

The biomass linear trend has an intercept value of ~140 and a slope of .47, showing a positive linear trend with a strong correlation of a very small p-value.

The renewable energy linear trend has an intercept of ~187 and a slope of ~.7, showing a positive linear trend with a strong correlation of a very small p-value. 

The hydro linear trend has an intercept of ~83 with a very strong, small p value,but a small, weak, negative linear trend with slope -0.01 with a p value that is still less than 0.05. It also has a weaker r squared value than the other two trends. 

### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r Need a date }
library(lubridate)
ym_date <- paste(raw_energy_data[,1])
ym_date <- ym(ym_date)  #function my from package lubridate
```

```{r Biomass Detrend}
biomass_beta0<- biomass_linear_trend$coefficients[1]
biomass_beta1 <- biomass_linear_trend$coefficients[2]


biomass_detrend <- energy_data[,1] - (biomass_beta0 + biomass_beta1*t)

df_biomass_detrend <-data.frame("date"=ym_date,
                        "observed"=energy_data[,1], 
                        "detrend" =biomass_detrend)
ggplot(df_biomass_detrend, aes(x=date))+
  geom_line(aes(y=observed),color="black")+
  geom_line(aes(y=detrend),color="blue")+
  labs(title="Biomass observed versus detrended")
```

This plot shows the upward linear trend removed, as well as centered the mean around zero. It also shows a greater dip around the year 2000.

```{r RE Detrend}
RE_beta0<- RE_linear_trend$coefficients[1]
RE_beta1 <- RE_linear_trend$coefficients[2]


RE_detrend <- energy_data[,2] - (RE_beta0 + RE_beta1*t)

df_RE_detrend <-data.frame("date"=ym_date,
                        "observed"=energy_data[,2], 
                        "detrend" =RE_detrend)
ggplot(df_RE_detrend, aes(x=date))+
  geom_line(aes(y=observed),color="black")+
  geom_line(aes(y=detrend),color="blue")+
  labs(title="RE observed versus detrended")
  
```

This plot removed the upward linear trend was removed and centered the mean at zero. It also shows a greater dip after 2000 relative to the rest of the data. 

```{r Hydro Detrend}
hydro_beta0<- hydro_linear_trend$coefficients[1]
hydro_beta1 <- hydro_linear_trend$coefficients[2]


hydro_detrend <- energy_data[,3] - (hydro_beta0 + hydro_beta1*t)

df_hydro_detrend <-data.frame("date"=ym_date,
                        "observed"=energy_data[,3], 
                        "detrend" =hydro_detrend)
ggplot(df_hydro_detrend, aes(x=date))+
  geom_line(aes(y=observed),color="black")+
  geom_line(aes(y=detrend),color="blue")+
  labs(title="Hydro observed versus detrended")
```

This plot did not change significantly beyond centering the mean at zero because there was very little trend component to this data.

### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. nut mot mandatory. Did the plots change? How?
```{r Make the detrend a TS}
ts_biomass_detrend <- ts(biomass_detrend,start=c(1973,1),frequency = 12)
ts_RE_detrend <- ts(RE_detrend,start=c(1973,1),frequency = 12)
ts_hydro_detrend <- ts(hydro_detrend,start=c(1973,1),frequency = 12)
```

```{r ACF detrended}
biomass_detrend_acf <- ggAcf(ts_biomass_detrend,lag.max=40)+
  labs(title="Biomass Detrended ACF")
#biomass_detrend_acf
hydro_detrend_acf <- ggAcf(ts_hydro_detrend,lag.max=40)+
  labs(title="Hydro Detrended ACF")
#hydro_detrend_acf
RE_detrend_acf <- ggAcf(ts_RE_detrend,lag.max=40)+
  labs(title="RE Detrended ACF")
#RE_detrend_acf
```

```{r PACF Detrend}
biomass_detrend_pacf <- ggPacf(ts_biomass_detrend,lag.max=40)+
  labs(title="Biomass Detrended PACF")
#biomass_detrend_pacf
hydro_detrend_pacf <- ggPacf(ts_hydro_detrend,lag.max=40)+
  labs(title="Hydro Detrended PACF")
#hydro_detrend_pacf
RE_detrend_pacf <- ggPacf(ts_RE_detrend,lag.max=40)+
  labs(title="RE Detrended PACF")
#RE_detrend_pacf
```

```{r Plotted together to compare}
plot_grid(biomass_detrend_acf,biomass_detrend_pacf,biomass_acf,biomass_pacf,
          ncol=2,align="h")
plot_grid(hydro_detrend_acf,hydro_detrend_pacf ,hydro_acf,hydro_pacf,
          ncol=2,align="h")
plot_grid(RE_detrend_acf,RE_detrend_pacf,RE_acf,RE_pacf,
          ncol=2,align="h")
```
The detrended biomass ACF shows much lower probability and connections over the periods than the detrended ACF, however the PACF shows some stronger probability on the detrended period and than on the original PACF. 

The detrended hydro ACF and PACF looks very similar to the original hydro ACF and PACF because there was very little trend to remove ot begin with, so this did not make a huge impact. 

The detrended RE ACF and PACF demonstrates a similar change to the detrended biomass figures. 

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Just by looking at the time series and the acf plots, do the series seem to have a seasonal trend? No need to run any code to answer your question. Just type in you answer below.

Just looking at the plots from Q1, hydro appears to have a strong seasonal trend. Biomass and RE appear to have what could be a weaker seasonal trend with their oscillating values at a somewhat regular period, with renewable energy appear to have more regular variations than biomass. 

### Q7
Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) the two time series. Ask R to print the summary of the regression. Interpret the regression output. From the results which series have a seasonal trend? Do the results match you answer to Q6?

```{r Biomass Seasonal Means}

#Use seasonal means model

#Create the seasonal dummies
biomass_dummies <- seasonaldummy(ts_energy_data[,1])  


#Fit a linear model to the seasonal dummies
biomass_seas_means_model <- lm(ts_energy_data[,1]~biomass_dummies)
summary(biomass_seas_means_model)


#Store regression coefficients
biomass_beta_int <- biomass_seas_means_model$coefficients[1]
biomass_beta_coeff <- biomass_seas_means_model$coefficients[2:12]

```
The biomass coefficients demonstrate there is a seasonal component, but this is a weak driver of the values in the model as the r-squared value is very low. 
```{r RE Seasonal Means}

#Use seasonal means model

#Create the seasonal dummies
RE_dummies <- seasonaldummy(ts_energy_data[,2])  


#Fit a linear model to the seasonal dummies
RE_seas_means_model <- lm(ts_energy_data[,2]~RE_dummies)
summary(RE_seas_means_model)


#Store regression coefficients
RE_beta_int <- RE_seas_means_model$coefficients[1]
RE_beta_coeff <- RE_seas_means_model$coefficients[2:12]

```
The coefficients demonstrate that there are seasonal components that affect the renewable energy as well, but these are very week as noted by their low R squared value. 

```{r Hydro Seasonal Means}
#Use seasonal means model

#Create the seasonal dummies
hydro_dummies <- seasonaldummy(ts_energy_data[,3])  


#Fit a linear model to the seasonal dummies
hydro_seas_means_model <- lm(ts_energy_data[,3]~hydro_dummies)
summary(hydro_seas_means_model)


#Store regression coefficients
hydro_beta_int <- hydro_seas_means_model$coefficients[1]
hydro_beta_coeff <- hydro_seas_means_model$coefficients[2:12]
```

The coefficients indicate a very strong seasonalcomponent to the data, with increased values in the summer and decreased values in the fall and winter months.These coefficients have a strong R square value and all have strong p values as well. This series has the most significant seasonal component of the three. 
### Q8
Use the regression coefficients from Q7 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

```{r Biomass seasonal decomp plots}
#compute seasonal component
biomass_seas_comp <- array(0,n_obs)
for(i in 1:n_obs){
  biomass_seas_comp[i] <- (biomass_beta_int+biomass_beta_coeff 
                           %*% biomass_dummies[i,])
}
#Viewing seasonal and original
ggplot(energy_data, aes(x=ym_date, y=energy_data[,1])) +
            geom_line(color="blue") +
            ylab(paste0(colnames(energy_data)[1])) +
            geom_line(aes(y=biomass_seas_comp), col="red")


#Removing seasonal component
deseason_biomass_data <- energy_data[,1]-biomass_seas_comp

#Plotted together
ggplot(energy_data, aes(x=ym_date, y=energy_data[,1])) +
            geom_line(color="blue") +
            ylab(paste0(colnames(energy_data)[1])) +
            geom_line(aes(y=deseason_biomass_data), col="green")

```

Biomass appears to have a very small seasonal trend that was removed through this process that smoothed out the regular variations year to year. De-seasoning the data also centered the mean at zero. 

```{r RE seasonal decomp plots}
#compute seasonal component
RE_seas_comp <- array(0,n_obs)
for(i in 1:n_obs){
  RE_seas_comp[i] <- (RE_beta_int+RE_beta_coeff %*% RE_dummies[i,])
}
#Viewing seasonal and original
ggplot(energy_data, aes(x=ym_date, y=energy_data[,2])) +
            geom_line(color="blue") +
            ylab(paste0(colnames(energy_data)[2])) +
            geom_line(aes(y=RE_seas_comp), col="red")


#Removing seasonal component
deseason_RE_data <- energy_data[,2]-RE_seas_comp

#Plotted together
ggplot(energy_data, aes(x=ym_date, y=energy_data[,2])) +
            geom_line(color="blue") +
            ylab(paste0(colnames(energy_data)[2])) +
            geom_line(aes(y=deseason_RE_data), col="green")
```

RE appears to have a very small seasonal trend that was removed through this process that smoothed out the regular variations year to year. De-seasoning the data also centered the mean at zero.
```{r Hydro Seasonal Decom Plots}
#compute seasonal component
hydro_seas_comp <- array(0,n_obs)
for(i in 1:n_obs){
  hydro_seas_comp[i] <- (hydro_beta_int+hydro_beta_coeff %*% hydro_dummies[i,])
}
#Viewing seasonal and original
ggplot(energy_data, aes(x=ym_date, y=energy_data[,3])) +
            geom_line(color="blue") +
            ylab(paste0(colnames(energy_data)[3])) +
            geom_line(aes(y=hydro_seas_comp), col="red")


#Removing seasonal component
deseason_hydro_data <- energy_data[,3]-hydro_seas_comp

#Plotted together
ggplot(energy_data, aes(x=ym_date, y=energy_data[,3])) +
            geom_line(color="blue") +
            ylab(paste0(colnames(energy_data)[3])) +
            geom_line(aes(y=deseason_hydro_data), col="green")
```
Deseasoning the hydro series smoothed out the regular yearly peaks and vallys in the hydro data. It also centered the mean at 0. This plot had the greatest change. 

### Q9
Plot ACF and PACF for the deseason series and compare with the plots from Q1. You may use plot_grid() again to get them side by side but not mandatory. Did the plots change? How?
```{r Make the deseason a TS}
ts_biomass_deseason <- ts(deseason_biomass_data,start=c(1973,1),frequency = 12)
ts_RE_deseason <- ts(deseason_RE_data,start=c(1973,1),frequency = 12)
ts_hydro_deseason <- ts(deseason_hydro_data,start=c(1973,1),frequency = 12)
```

```{r ACF Deseason}
biomass_deseason_acf <- ggAcf(ts_biomass_deseason,lag.max=40)+
  labs(title="Biomass Deseasoned ACF")
#biomass_season_acf
hydro_deseason_acf <- ggAcf(ts_hydro_deseason,lag.max=40)+
  labs(title="Hydro Deseasoned ACF")
#hydro_season_acf
RE_deseason_acf <- ggAcf(ts_RE_deseason,lag.max=40)+
  labs(title="RE Deseasoned ACF")
#RE_season_acf

```

```{r PACF Deseason}
biomass_deseason_pacf <- ggPacf(ts_biomass_deseason,lag.max=40)+
  labs(title="Biomass Deseasoned PACF")
#biomass_season_pacf
hydro_deseason_pacf <- ggPacf(ts_hydro_deseason,lag.max=40)+
  labs(title="Hydro Deseasoned PACF")
#hydro_season_pacf
RE_deseason_pacf <- ggPacf(ts_RE_deseason,lag.max=40)+
  labs(title="RE Deseasoned PACF")
#RE_season_pacf

```

```{r Plotted together to compare}
plot_grid(biomass_deseason_acf,biomass_deseason_pacf,biomass_acf,biomass_pacf,
          ncol=2,align="h")
plot_grid(hydro_deseason_acf,hydro_deseason_pacf ,hydro_acf,hydro_pacf,
          ncol=2,align="h")
plot_grid(RE_deseason_acf,RE_deseason_pacf,RE_acf,RE_pacf,
          ncol=2,align="h")
```

The biomass and renewable energy PACF and ACFs look very similar to their deseasoned versions. However, the hydro deseasoned ACF shows far less correlation to previous data due to taking out the seasonal similarities,and the PACF has an even smaller statistical significance in probability the further along the lag goes. Taking out the seasonal component really helped the hydro data to be less dependent on many periods before it. 